---
title: "Validatetools"
subtitle: "Validatetools: Check and resolve contradictory rule sets" 
author: "Edwin de Jonge en Mark van der Loo"
date: "uRos2018 Tutorial Session, The Hague"
output:
  beamer_presentation:
    keep_tex: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE)
library(validatetools)
```

## {.plain}

\hspace*{-1cm}\includegraphics[width=0.9\paperwidth]{img/bad-data}

## Desirable data cleaning properties: 

- Reproducible data checks.
- Automate repetitive data checking (e.g. monthly/quarterly).
- Monitor data improvements / changes. 
- **How** do this systematically?

## {.plain}

\hspace*{-1cm}
\includegraphics[width=1\paperwidth]{img/rules.png}

## Data Cleaning philosophy

- **"Explicit is better than implicit"**.
- Data rules are solidified **domain knowledge**.
- Store these as **validation rules** and apply these when necessary.

### Advantages:

- Easy checking of rules: data validation.
- Data quality statistics: how often is each rule violated? 
- Allows for reasoning on rules: which variables are involved in errors? How do errors affect the resulting statistic?
- Simplifies rule changes and additions.

## Refresh: R package `validate`

With package `validate` you can formulate explicit rules that data must conform to:

```{r, eval=FALSE}
library(validate)
check_that( data.frame(age=160, job = "no", income = 3000), 
  age >= 0, 
  age < 150,
  job %in% c("yes", "no"),
  if (job == "yes") age >= 16,
  if (income > 0) job == "yes"
)
```

## Rules (2)

A lot of datacleaning packages are using validate rules to facilitate their work.

- `validate`: validation **checks** and data **quality stats** on data. 
- `errorlocate`: to find **errors** in variables (in stead of records)
- `rspa`: data **correction** under data constraints
- `deductive`: deductive **correction**
- `dcmodify`: deterministic **correction** and **imputation**.

## Growing pains

- using explicit rules is great, but when succesful create new and 
unforeseen issues.

### Issues:

- Many variables.
- Many rules, checks or constraints on the data.
- Many sub-domains with specialized rules.
- Many persons working on same rule set.

## Assigment 1 (5 min.)

- Collect in small groups:
  - What is maximum columns in a dataset you encountered in your office?
  - Can you give an indication of the maximum number of data validity rules that are checked in production process ? 
  - How many persons are involved in checking and maintaining the rules?

## Issues:

At CBS:

- Datasets with > 100 columns are common.
- Some systems have 100s of rules.
- Often multiple persons work on rule set.

Most of these issues are not technical, but **organisational** and **cognitive**.

- Does anyone has a clear oversight on a large rule dataset? 
- If your co-worker adds a rule, this (may) interfere with the other rules.

## Why-o-why `validatetools`?

- We have package `validate`, what is the need?

### Because we'd like to...

- clean up rule sets ( kind of meta-cleaning...).
- detect and resolve problems with rules:

    - Detect unintended rule **interactions**.
    - Detect **conflicting** rules.
    - Remove **redundant** rules.
    - **Substitute** values and **simplify** rules.

- check the rule set using formal logic (**without any data!**).
- solve these kind of fun problems :-)

## Detect rule interactions

- The rules form a consistent system of constraints.

- A combination of rules may _overconstrain_ a variable

- One simple option is look at the boundary of allowed values for 
each variable.


## Assignment Check boundaries

### 

1) What are the allowed values for `age` and `income`?
```{r, eval = FALSE}
library(validatetools)
rules <- validator( age >= 18
                  , if (job == TRUE) age <= 70
                  , if (income > 0) job == TRUE
                  , income >= 0
                  )
```
2) Check this with `validatetools::detect_boundary_num`.

## 

```{r, eval = FALSE}
library(validatetools)
rules <- validator( age >= 18
                  , if (job == TRUE) age <= 70
                  , if (income > 0) job == TRUE
                  , income >= 0
                  )
detect_boundary_num(rules)
```

## Rule interactions:

- boundary check is ok, may does not check for forbidden intervals.
- when variable can only have one value, it is fixed.
- extreme case is when allowed range for a variable is empty: infeasibility

## Problem: infeasibility

### Problem

One or more rules in conflict: all data incorrect, because always one of the rules will be violated! (_and yes that happens when rule sets are large ..._).

### `validatetools` checks for feasiblity

```{r}
library(validatetools)
rules <- validator( is_adult = age >=21
                  , is_child = age < 18
                  )
is_infeasible(rules)
```

## {.plain}

\hspace*{-3cm}
\begin{center}
\includegraphics[height=1\paperheight]{img/keepcalm-and-resolve.png}
\end{center}

## Conflict, and now?

```{r}
rules <- validator( is_adult = age >=21
                  , is_child = age < 18
                  )
# Find out which rule would remove the conflict
detect_infeasible_rules(rules)
# And its conflicting rule(s)
is_contradicted_by(rules, "is_adult")
```

- One of these rules needs to be removed
- Which one? Depends on human assessment...

## Assignment Find the conflicting rules

a) Open the file "infeasible_rules.txt" (e.g. `file.edit("infeasible_rules.txt")`).
Can you see which records are in conflict?

b) Find which two rule(s) are causing the infeasibility in file "infeasible_rules.txt".

```{r}
rules <- validator(.file = "infeasible_rules.txt")
is_infeasible(rules)
# do your thing...
```

## Detecting and removing redundant rules

- Often rule set contain redudent rules. 
- This may seem not a problem, however:

    - it complicates the rule set
    - it makes automatic checking a lot more problematic.

## Detecting and removing redundant rules

Rule $r_1$ may imply $r_2$, so $r_2$ can be removed.

```{r}
rules <- validator( r1 = age >= 18
                  , r2 = age >= 12
                  )
detect_redundancy(rules)
remove_redundancy(rules)
```

## Value substitution

In complex statistics, many rules are specific for sub domains/sub groups

- This can be mitigated by splitting the rule sets in different pieces
- But can also be handled by simplifying the rule set for each subdomain:

- Fill in a value into a variable (making it a constant) and simplify the remaining rules.

## Value substitution

```{r}
rules <- validator( r1 = if (gender == "male") weight > 50
                  , r2 = gender %in% c("male", "female")
                  )

substitute_values(rules, gender = "male")
```

## Assignment: Can you simplify this one?

a) 

```{r}
validator( if (income > 0) age >= 16
         , age < 12
        )
```
b) Use `simplify_conditional` to let validatetools do it.

## 

A bit more complex reasoning, but still classical logic:

```{r}
rules <- validator( r1 = if (income > 0) age >= 16
                  , r2 = age < 12
                  )
# age > 16 is always FALSE so r1 can be simplified
simplify_conditional(rules)
```

## Assigment: all together now!

`simplify_rules` applies all simplification methods to the rule set.

a) If we know that job must be "yes", can you see how this rule set can  be simplified?

```{r, eval = FALSE}
rules <- validator( r1 = job %in% c("yes", "no")
                  , r2 = if (job == "yes") income > 0
                  , r3 = if (age < 16) income == 0
                  )
```
b) Apply `simplify_rules(rules, job = "yes")`
c) Can you do the same using the other simplifying functions?


## 

```{r}
rules <- validator( r1 = job %in% c("yes", "no")
                  , r2 = if (job == "yes") income > 0
                  , r3 = if (age < 16) income == 0
                  )
simplify_rules(rules, job = "yes")
```

## How does it work?

`validatetools`:

- reformulates rules into formal logic form.

- translates them into a mixed integer program for each of the problems.

### Rule types

- _linear_ restrictions
- _categorical_ restrictions
- _if_ statements with linear and categorical restrictions

### If statement is Modus ponens:

$$
  \begin{array}{ll}
                  & \textsf{if } P \textsf{ then } Q \\
  \Leftrightarrow & P \implies Q \\
  \Leftrightarrow & \lnot P \lor Q
  \end{array}
$$

## Example

```{r}
rules <- validator(
  example = if (job == "yes") income > 0
)
```
$$
r_{\textrm{example}}(x) = \textrm{job} \not \in \textrm{"yes"} \lor  \textrm{income} > 0  
$$
```{r}
print(rules)
```

## Addendum

## Formal logic

### Rule set $S$

\newcommand{\la}[1]{\boldsymbol{#1}}

A validation rule set $S$ is a conjunction of rules $r_i$, which applied on record $\la{x}$ returns `TRUE` (valid) or `FALSE` (invalid)

$$
S(\la{x}) = r_1(\la{x}) \land \cdots \land r_n(\la{x})
$$

### Note 

- a record has to comply to each rule $r_i$.

- it is thinkable that two or more $r_i$ are in conflict, making each record invalid.

## Formal logic (2)

### Rule $r_i(x)$

A rule a disjunction of atomic clauses:

$$
r_i(x) = \bigvee_j C_i^j(x)
$$
with: 

$$
C_i^j(\la{x}) = \left\{
  \begin{array}{l}
 \la{a}^T\la{x} \leq b \\
  \la{a}^T\la{x} = b \\
  x_j \in F_{ij}  \textrm{with } F_{ij} \subseteq D_j \\
  x_j \not\in F_{ij} \textrm{with } F_{ij} \subseteq D_j \\
  \end{array}
\right.
$$


## Mixed Integer Programming

Each rule set problem can be translated into a mip problem, which can be 
readily solved using a mip solver. 

`validatetools` uses `lpSolveApi`.

$$
\begin{array}{r}
  \textrm{Minimize } f(\mathbf{x}) = 0; \\
  \textrm{s.t. }\mathbf{Rx} \leq \mathbf{d} \\
\end{array}
$$
with $\la{R}$ and $\la{d}$ the rule definitions and $f(\la{x})$ is the specific problem that is solved.

